Computational Prediction of Boundaries of 3D Genomic Domains in Class Imbalance Settings

### 2020-05-29 Manuscript/manuscript.Rmd
editted intro

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed important 3-dimensional (3D) spatial constructs within the human genome such as loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. Among these, TADs represent regions on the linear genome, at megabase scale, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics like gene regulation and cell differentiation, an example being the mechanism by which enhancers loop to come into contact with distal promoters to initiate transcription within these domains [@Lupianez:2015aa; @Franke:2016aa]. Likewise, disruption of TAD boundaries has been shown to lead to developmental disorders [@lupianez2016breaking], cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@Sun:2018aa; @Meaburn:2007aa]. Therefore, identifying the location of TAD boundaries remains a top priority in our goal to understand the functionality of the human genome.

A key factor in locating TAD boundaries is "resolution". Due to the high-dimensional nature of Hi-C data, genome binning has been a common strategy for dealing with a low signal-to-noise ratio problem [@li2019comparison]. Resolution, therefore, determines the bin sizes of the linear genome used to enumerate contacts caused by chromatin folding. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define sub-domains and the boundaries that demarcate them [@fudenberg2016formation]. Furthermore, even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-Sequencing experiments used to annotate the genome, which ranges in the tens of base pairs [@mendoza2009genome]. This resolution gap between 1D and 3D genomic maps can make it difficult to disentangle the role of genomic factors related to TAD boundaries, thereby limiting our understanding of epigenomic regulation and its potential link to chromatin architecture. We hypothesize that, by leveraging genomic annotations--defined by high resolution ChIP-Seq data--known to be enriched at TAD boundaries, we may be able more to precisely predict the locations of TAD boundaries at higher resolutions.

Several methods have been developed to predict TAD boundary location using genomic annotations. However, most ignore key characteristics of 3D genomic data that are detrimental to both model performance and accurate boundary identification. A method developed by Mourad et al [@mourad2016computational], called *HiCFeat*, used the percentage of overlap between several ChIP-seq defined transcription factor binding site (TFBS) regions and 1 kb genomic bins to build an L1-regularized multiple logistic regression model to predict TAD boundary regions. However, such high level of resolution is likely to introduce heavily imbalanced classes created from the proportionally much smaller number of TAD boundary regions compared to non-TAD boundary regions throughout the linear genome. *HiCFeat* did not the address class imbalance. Instead, model performance was evaluated using area under the receiver operating characteristic curve (AUROC) which is known to be insensitive to class imbalance, creating artificially inflated values influenced by the majority class [@prati2011survey; @wei2013role]. <!--A more appropriate evaluation tool to consider would be a composite metric such as Matthews Correlation Coefficient (MCC), which penalizes for both false positives and false negatives.--> Furthermore, by only considering the percentage of overlap in defining the feature space, *HiCFeat* is limited in its granularity given that many regions on the linear genome will not be overlapped with any genomic annotations. <!--This, likely being further exacerbated by the class imbalance.--> Two additional studies were proposed in 2015 and 2017 respectively, one using histone modifications in a Bayesian Additive Regression Trees (BART) model [@huang2015predicting] and the other using combined sets of TFBS, DNase I hypersensitive sites, and histone modifications together in an L2-based regularized linear model [@hong2017computational]. The BART model was built on relatively high resolution Hi-C data at 5 kb, while the L2-regularized model used 200 kb resolution Hi-C data. Both methods addressed class imbalance by performing random under-sampling (RUS), and used the elemental read count that appeared in each bin to describe the relationship between ChIP-seq annotations and genomic regions. Firstly, given the large disparity between the number of TAD vs non-TAD boundary regions, performing random under-sampling alone is likely to be unstable and introduce bias. Also, it is unclear how other re-sampling solutions compare to simple random under-sampling such as random over-sampling or a weighted combination of both random under- and over-sampling together. Secondly, using read count overlap as the feature space suffers from similar limitations as the percentage of overlap. Moreover, it is unclear how either of these feature engineering steps compare to enumerating the distance in base pairs between ChIP-seq annotations and genomic bins in TAD boundary prediction, which offer a more spatial measure of association.

Given the many pitfalls encountered by previous methods we set out to establish a unified approach toward TAD boundary prediction. In this study we first developed a computational framework aimed at determining the set of data level characteristics necessary for optimal TAD boundary prediction. We assessed the Matthews Correlation Coefficient (MCC) of 10-fold cross-validated random forest classifiers with different combinations of model inputs consisting of vary data resolutions, re-sampling techniques, and feature engineering steps. We showed that 10 kb resolution data, combined with both random over- and under-sampling, and a feature space composed of distances between ChIP-seq annotations and genomic bins yielded the best model performances for TAD boundary prediction. We showed that our optimally built model was able to identify known architectural proteins CTCF, RAD21, SMC3, and ZNF143 as exhibiting high predictive importance, confirming their biological functionality near TAD boundaries. We then aimed to use our optimally built model to more precisely predict the location of TAD boundaries at base pair resolution. We showed that, compared to TAD boundaries identified by a TAD caller [@durand2016juicebox], these top ranked transcription factors were much more enriched near the predicted TAD boundaries identified by our model indicating much more precise locations throughout the linear genome. This improved precision in TAD boundary location can provide insight into the role that epigenomics has on the spatial architecture of the human genome.

<!--
Extensive research using high resolution maps (ranging from 750 bp to 1 mb) have probed much of the aspects of 3D chromatin architecture identifying several epigenomic modifications associated with TAD boundaries [@rao20143d; @bonev2017multiscale]. For instance, it is known that the CCCTC-binding factor (CTCF) is enriched at TAD boundaries acting as an insulator interfering with enhancer-promotor interactions [@ghirlando2016ctcf; @song2017ctcf; @ong2014ctcf]. Additional transcription factors including SMC1, SMC3, and RAD21 have also been shown to colocalize near TAD boundaries forming a cohesin ring under the proposed loop extrusion model, whereby loops are formed during interphase [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. 
-->

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@fudenberg2016formation; @alipour2012self].

@@

Our computational framework alone yielded many interesting results. First, we showed that RF models built using *distance*-type predictors out performed models built on previously published feature engineering techniques including overlap counts and overlap percents. Moreover, we were further able to optimize our RF model by introducing random under-sampling, a technique to correct for the severe class imbalance between TAD(loop) and non-TAD(loop) boundary regions. We showed that in our optimized RF, transcription factor binding sites provided an overwhelming amount of signal in the model, indeed out-performing histone modification- and BroadHMM-specific models, contradicting a previous study that found histone marks alone were effective in predicting TAD boundaries [@huang2015predicting]. Likewise, our RF model built using *distance*-type predictors was consistently able to identify the group of the top known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 as being highly predictive of TAD(loop) boundary regions across both cell lines. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. This suggested that random forests, when tuned and feature engineered correctly, are highly effective in accurately predicting boundary regions, contradicting a previous study that found simple multiple linear regression out performed random forests in a TAD boundary prediction setting using both human and drosophila Hi-C data [@mourad2016computational]. Plausible reasons for their findings include lack of class balancing prior to modeling and usage of rudimentary feature engineering procedures including overlap counts and overlap percents.

>>>>

The advent of chromosome conformation capture (3C) sequencing technologies, and its successor Hi-C, have revealed a hierarchy of important 3-dimensional (3D) spatial constructs within the human genome such as chromatin loops [@rao20143d], Topologically Associated Domains (TADs) [@nora2012spatial;@dixon2012topological], and A/B compartments [@lieberman2009comprehensive]. At kilobase scale, gene promotors are often connected via distal regulatory elements, such as enhancers, through chromatin loops, thereby regulating gene expression [@Lupianez:2015aa; @Franke:2016aa]. Further out, at the megabase scale, TADs represent regions on the linear genome, that are highly self-interacting. Emerging evidence has linked TADs to critical roles in cell dynamics and cell differentiation. It has been shown that TADs themselves are highly conserved across species and cell lines [@dixon2012topological;@krefting2018evolutionary]. Likewise, disruption of the boundaries that demarcate TADs and loops have been shown to lead to cancer [@hnisz2016activation; @taberlay2016three] and other disorders [@lupianez2016breaking;@Sun:2018aa;@Meaburn:2007aa]. Therefore, precisely identifying the location of TAD and loop boundaries remains a top priority in our goal to fully understand the functionality of the human genome.

Several methods have been proposed to identify the genomic coordinates of TAD boundaries throughout the linear genome [@dixon2012topological;@rao20143d;@sexton2012three;@chen2016novo;@filippova2014identification;@shin2016topdom;@weinreb2016identification], less so for chromatin loops [@salameh2019supervised;@kai2018predicting]. However, a key limitation among most of these "TAD(loop)-callers" is that they are heavily reliant on Hi-C data *resolution*. Resolution refers to the sizes of bins used to segment the linear genome as a way to deal with low signal-to-noise incurred by the high dimensional nature of Hi-C data [@li2019comparison]. Lower resolutions contribute to larger bin sizes, which lead to variability in the location of TAD boundaries. Due to high sequencing costs, most available Hi-C datasets have relatively low resolution, ranging from 25 kb to 100 kb. Because increasing the resolution of Hi-C data requires a quadratic increase in the total sequencing depth, obtaining high resolution remains difficult [@schmitt2016genome]. While these low resolution Hi-C datasets can be used to probe higher order genomic patterns such as A/B compartments, more refined resolutions are necessary to accurately define chromatin loops and TADs [@fudenberg2016formation].

Another limitation among TAD(loop)-callers is their disregard for the fundamental roles of functional genomic elements on chromatin folding. It has been shown that particular genomic elements are enriched at TAD(loop) boundaries. Namely, the insulator binding protein, CTCF, and additional cofactors such as SMC3 and RAD21 have been hypothesized to comprise components of the loop extrusion model, whereby DNA is extruded through cohesin rings forming loops [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin]. Additionally, distinct patterns of histone modifications have also been shown to be featured at boundaries, including H3K36 timethylation (H3K36me3) [@lieberman2009comprehensive;@dixon2012topological;@libbrecht2015joint]. These elements are annotated on the genome using chromatin immunoprecipitation followed by high-throughput DNA sequencing (ChIP-seq). Even at its highest (750 bp [@bonev2017multiscale]), Hi-C sequencing resolution still remains well below the resolution of ChIP-seq experiments, which ranges in the tens of base pairs [@mendoza2009genome]. Therefore, leveraging such elements in a supervised learning framework lends itself to the possibility of more precisely predicting the locations of TAD (or loop) boundaries at higher resolutions.

???stopped here

@@

We sought to further optimize our TAD prediction model by performing recursive feature elimination to avoid overfitting and select only the most influential TFBS across all chromosomes. We were able to obtain near-optimal performance using approximately eight TFBS (Figure 4B; Supplementary Figures 4B). However, given that we trained our model on chromosome-specific data, the most significant annotations were found to vary for each chromosome. To determine transcription factors most important for boundary prediction, we clustered the predictive importances of the top eight significant TFs across chromosomes. We found that, similar to other studies [@sanborn2015chromatin; @weintraub2017yy1; @gonzalez2017znf143], CTCF, RAD21, SMC3, and ZNF143 transcription factors were most predictive of TAD and loop boundaries for all chromosomes (Figure 4C; Supplementary Figures 4C). We further optimized our model by only considering these top informative TFBS when building the random forest, thereby decreasing computational burden while maintaining high predictive performance. In summary, our framework was able to yield the top molecular drivers of the 3D architecture of the human genome, components of the popular loop extrusion model [@hansen2018recent;@fudenberg2016formation;@alipour2012self;@sanborn2015chromatin].

@@

Our computational framework alone yielded many interesting results. First, we showed that RF models built using *distance*-type predictors out performed models built on previously published feature engineering techniques including overlap counts and overlap percents. Moreover, we were further able to optimize our RF model by introducing random under-sampling, a technique to correct for the severe class imbalance between TAD(loop) and non-TAD(loop) boundary regions. We showed that in our optimized RF, transcription factor binding sites provided an overwhelming amount of signal in the model, indeed out-performing histone modification- and BroadHMM-specific models, contradicting a previous study that found histone marks alone were effective in predicting TAD boundaries [@huang2015predicting]. Likewise, our RF model built using *distance*-type predictors was consistently able to identify the group of the top known molecular drivers of 3D chromatin including CTCF, RAD21, SMC3, and ZNF143 as being highly predictive of TAD(loop) boundary regions across both cell lines. These are known components of the loop extrusion model, an established theory for how loops are made by a ring-shaped adenosine triphosphatase complex called cohesin [@hansen2018recent;@sanborn2015chromatin; @fudenberg2016formation; @alipour2012self]. This suggested that random forests, when tuned and feature engineered correctly, are highly effective in accurately predicting boundary regions, contradicting a previous study that found simple multiple linear regression out performed random forests in a TAD boundary prediction setting using both human and drosophila Hi-C data [@mourad2016computational]. Plausible reasons for their findings include lack of class balancing prior to modeling and usage of rudimentary feature engineering procedures including overlap counts and overlap percents.
